---
layout: post
title:  Re-coding categorical variables in R
description: coding categorical variables in R with the factor and ifelse command
share: true
tags:
    - R
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

While I have done data analysis almost exclusively in R for most of my career, until recently I have done a lot of my data cleaning in Stata. I have even told people in the past that I though Stata was better at data cleaning than R. However, when I switched my graduate statistics class from Stata to R last year, I decided that I needed to teach my students how to clean data in R. True to the old saying, there is no better way to learn than teach. In this case, I realized that my attitude toward data cleaning in R was more about my lack of familiarity with some nice utilities in R than it was about R itself. 

In this post, I wanted to use some code from one of my current projects to outline how to efficiently re-code categorical variables in R. I will first outline how I used to do it and then demonstrate much simpler code to achieve the same effect. I am not using any additional packages here, including the [tidyverse](https://www.tidyverse.org/). All code is base R.

## The Data

The data I am using come from a [data extract](https://gssdataexplorer.norc.org/projects/40124) of the General Social Survey (GSS). I am particularly interested in the variable `PRAY` in this dataset. Yes, the GSS capitalizes all of its variable names because REASONS. You can access the CSV data here. 

```{r read_data}
gss <- read.csv("gss_relig.csv")
table(gss$PRAY)
```

The default data came in numeric format. The [online codebook for the GSS](https://gssdataexplorer.norc.org/variables/315/vshow) tells me that these numeric codes correspond to the following categories. 

Numeric Code | Category
-------------|---------------------------------------------
0            | Not applicable (wasn't asked in that year)
1            | Several times a day
2            | Once a day
3            | Several times a week
4            | Once a week
5            | Lt once a week
6            | Never
8            | Don't know
9            | No answer

Ultimately, I want to turn these numeric codes into a proper factor variable. There are several issues to consider before proceeding:

1. There are multiple missing value codes to enter (0,8,9)
2. The numbering is the reverse of how I want the ordinal variable coded. I want higher values to indicate higher frequency of prayer.
3. Most importantly, the category of "Never" was not given as an explicit option until 2004, so I can't compare data across time very well with that category. Therefore, I want to collapse the "Never" and "Lt once a week" category together. 

## My Old Approach

My old approach to handling this sort of re-code problem would have been to apply what I call the **bracket and boolean** approach. Here is what that code would look like: 

```{r old_approach}
gss$pray <- NA
gss$pray[gss$PRAY==5 | gss$PRAY==6] <- "lt once a week"
gss$pray[gss$PRAY==4] <- "once a week"
gss$pray[gss$PRAY==3] <- "several times a week"
gss$pray[gss$PRAY==2] <- "once a day"
gss$pray[gss$PRAY==1] <- "several times a day"
gss$pray <- factor(gss$pray)
gss$pray <- relevel(gss$pray, "several times a day")
gss$pray <- relevel(gss$pray, "once a day")
gss$pray <- relevel(gss$pray, "several times a week")
gss$pray <- relevel(gss$pray, "once a week")
gss$pray <- relevel(gss$pray, "lt once a week")
table(gss$pray, gss$PRAY, exclude=NULL)
```

First, I created a new variable full of missing values. I then used indexing and boolean statements to replace those missing values with character strings of the correct categories. I then had turned those character string into a factor with the `factor` command. However, by default, that command will order my categories alphabetically, so I then ran *five* relevel commands to get the ordering correct. I then run a `table` command with the `exclude=NULL` option to include missing values to [check myself before I wreck myself](https://www.youtube.com/watch?v=bCY9L3Xidoo). It works, but I am not proud of that code. 

Had I bothered to read the help file for `factor`, I would have realized that I could actually specify ordering of the categories with the `levels` argument, so I could have simplified my code a little bit by removing the `factor` and `relevel` commands and replacing them with the following:

```{r better_factor_command, eval=FALSE}
gss$pray <- factor(gss$pray,
                   levels=c("lt once a week","once a week",
                            "several times a week","once a day",
                            "several times a day"),
                   ordered=TRUE)
```

The `ordered` option is particularly useful here for running greater than or equal to booleans on factor variables, as I will demonstrate below. Using this code will improve my initial code considerably, but I can still do much better. 

## My New Approach

My new approach takes advantage of two functions. First, the `factor` function itself is very flexible and can often be used to recode categorical variables without resort to the **bracket and boolean** approach. In this case, I can take advantage of the `levels` and `labels` argument to directly code my original numeric variable into a categorical variable in one line. 

The second function that is essential to date cleaning in R is the `ifelse` command. This function takes a boolean statement and assigns the value in the second argument if TRUE, and the value in the third argument if FALSE. In this case, I can easily combine the 5 and 6 numeric values from the original variable into just the value of 5 with:

```{r ifelse_example, eval=FALSE}
ifelse(gss$PRAY==6, 5, gss$PRAY)
```

To do the entire re-coding in one line, I just need to wrap a `factor` function around that code as follows: 

```{r new_approach}
gss$pray <- factor(ifelse(gss$PRAY==6, 5, gss$PRAY), 
                   levels=5:1,
                   labels=c("lt once a week","once a week",
                            "several times a week","once a day",
                            "several times a day"), 
                   ordered = TRUE)
table(gss$pray, gss$PRAY, exclude=NULL)
```

Just like that, twelve lines of code become one. 

You can apply this same trick to data to re-code and collapse data that is already coded as a factor variable. You just need to use cascading `ifelse` command to collapse and use a character vector for the levels.

To demonstrate, I use my new `pray` variable to create a `pray.simple` variable that collapses the data as follows:

- "lt once a week"
- "weekly": "once a week" and "several times a week"
- "daily": "once a day" and "several times a day"

I can do this again in a single line of code by using *cascading* `ifelse` statements. When the result of an `ifelse` statement is FALSE, I use another `ifelse` statement to continue looking for categories that I need. 

```{r cascading_ifelse}
gss$pray.simple <- factor(ifelse(is.na(gss$pray), NA,
                                 ifelse(gss$pray>="once a day", "daily",
                                        ifelse(gss$pray>="once a week",
                                               "weekly", "lt once a week"))),
                          levels=c("lt once a week", "weekly", "daily"), 
                          ordered=TRUE)
table(gss$pray, gss$pray.simple, exclude=NULL)
```

The first `ifelse` command filters out the missing values. Then I can use `>=` booleans (because I used the `ordered=TRUE` on the factor variable) to partition my valid categories and collapse them into the three that I want. I then define the ordering of the categories by feeding a character vector into `levels`. The `table` command shows that everything is ended up where I expected it to be. 

Now that my variable is all coded properly, I can look at the trend over time in the frequency of prayer:

```{r trend, dpi=300}
p <- prop.table(table(gss$YEAR, gss$pray.simple),1)
matplot(as.numeric(rownames(p)), 100*p, type="l", lty=1, col=rainbow(3),
        lwd=2, las=1, ylab="percent", xlab="year", ylim=c(0,80),
        main="Trend in frequency of prayer in the US,\nGeneral Social Survey")
legend(1972, 80, legend=levels(gss$pray.simple), lty=1, col=rainbow(3),
       ncol=3)
```